---
title: "Corhort_analysis_in_R"
author: "Amanda Wu"
date: "2023-05-25"
output:
  html_document:
    df_print: paged
---

# Load libraries
```{r}
# install.packages("dplyr") # to install this package for joining data frames
library(dplyr)

# install.packages('ascii') # to install this package for joining matrix
library(ascii)

# install.packages("tidyr") # to install this package for data wrangling
library(tidyr)

# install.packages("tidyverse"), lubricate package is within tidyverse; lubricate is used for converting time zones
library(lubridate)

# The easiest way to get ggplot2 is to install the whole tidyverse: install.packages("tidyverse")
library(ggplot2)

# install.packages("gridExtra"); this package helps combine multiple ggplot together 
library(gridExtra)
```

# Read data files and examine the data structure 
```{r}
# define the directory 
pathname<-"C:/Users/Amanda/Desktop/jobs/job_prep_materials/Applying/Invitae/Cohort_Analysis/Cohort_Analysis/" # this is the directory where I stored the csv files (please modify according to your directory)

# read csv files 
customer.df<-read.csv(paste0(pathname, "customers.csv")) # customer 
order.df<-read.csv(paste0(pathname, "orders.csv")) # order 

# Look at the basics of customer.df
head(customer.df)
dim(customer.df) # 25716 rows of data 
# change column names 
colnames(customer.df)<-c("user_id", "sign_up_date")

# Look at the basics of order.df
head(order.df)
dim(order.df) # 27575 rows of data 
# change column names 
colnames(order.df)<-c("order_id", "order_number", "user_id", "order_date")
```

# Join the two data frames and do some data wrangling 
```{r}
# join the two data frame based on column user_id
join.df<-full_join(order.df,customer.df,  by = "user_id")
head(join.df) # notice that some rows have missing data, because the customer.df is a smaller subset within order.df

# delete rows where either sign_up_date and order_date is empty
delete.index<-unique(c(which(is.na(join.df$sign_up_date)),  which(is.na(join.df$order_date))))
join.complete.df<-join.df[-delete.index,]

# check the data frame
join.complete.df[1:20,]
tail(join.complete.df)
nrow(join.complete.df) # 17015 rows left

# check if all order date is later than sign up date
which(join.complete.df$order_date < join.complete.df$sign_up_date ) # yes, indeed all order date is later than sign up date

# check if some customers have multiple times of order
customer.ordertimes<-join.complete.df %>% group_by(user_id) %>% summarise(order_times = n())
hist(customer.ordertimes$order_times, breaks = 50) # yes, indeed some customers ordered more than 1 time 

# Convert UTC time to PDT 
for(columname in  c("order_date", "sign_up_date")){
join.complete.df[,columname]<-as.POSIXct(join.complete.df[,columname], tz="UTC") # define the current format as UTC
print(join.complete.df[,columname][1:6]) # print original time in UTC
join.complete.df[, columname]<-with_tz(join.complete.df[,columname], "America/Los_Angeles") # convert from UTC to PDT(Pacific Daylight Time) time zone  
print(join.complete.df[,columname][1:6]) # print to visually confirm that the conversion is correct
}

# calculate the days between sign up date and order date
join.complete.df$days_after_signup<-as.numeric(floor((join.complete.df$order_date-join.complete.df$sign_up_date)/1440 ))
head(join.complete.df)

# calculate the week number between sign up date and order date
join.complete.df$weeks_after_signup<-as.numeric(floor(join.complete.df$days_after_signup/7)+1)
```

# Do cohort analysis for total users (including repetitive buyers)
```{r}
# first group the customer by the week they signed up
join.complete.df$CohortGroup<-cut(join.complete.df$sign_up_date, breaks="7 days", labels=FALSE)
# how many bins should there be?
as.numeric((max(join.complete.df$sign_up_date)-min(join.complete.df$sign_up_date))/7) # there should be 27 bins
# check if the time bins are correct 
range(join.complete.df$sign_up_bin)
join.complete.df[1:20,]
tail(join.complete.df)

# create reference data frame of total users for each cohort group
base_cohort_df = join.complete.df %>% group_by(CohortGroup) %>% summarise(TotalUsers = n_distinct(user_id), cohortname= paste(substr(min(sign_up_date),1,10),"to",substr(max(sign_up_date),1,10)))
head(base_cohort_df)

# create purchase activity data frame
activity_cohort_df = join.complete.df %>% group_by(CohortGroup, weeks_after_signup) %>% summarise( BuyingUsers = n_distinct(user_id))
head(activity_cohort_df)

# join activity_cohort_df and base_cohort_df
user_cohort_df = inner_join(activity_cohort_df, base_cohort_df, 
                  by = 'CohortGroup')

# Cohort table by the counts of users
cohort_retention_df = user_cohort_df %>% select(CohortGroup,  cohortname, TotalUsers, weeks_after_signup, BuyingUsers) %>%spread(weeks_after_signup, BuyingUsers)
head(cohort_retention_df)
# change the column names 
for (colno in colnames(cohort_retention_df)[-1:-3]){
  colnames(cohort_retention_df)[as.numeric(colno)+3]<-paste0("Days",as.numeric(colno)*7-7,"to",as.numeric(colno)*7-1)
}
head(cohort_retention_df)

# the percentage version of the dataframe 
cohort_retention.perc<-sapply(cohort_retention_df[,-1:-3], function(x) paste0(round(x/cohort_retention_df$TotalUsers*100,2),"% orderers"))
cohort_retention.perc<-gsub( "NA% orderers", "",cohort_retention.perc)

# bind the count user matrix and percentage user matrix together
cohort_retention_df.perc<-paste.matrix(cohort_retention.perc, "(", cohort_retention_df[,-1:-3],")")
colnames(cohort_retention_df.perc)<-colnames(cohort_retention.perc)
cohort_retention_df.perc<-cbind(cohort_retention_df[,1:3],cohort_retention_df.perc)
head(cohort_retention_df.perc)
```

# Do cohort analysis for first time orderers
```{r}
# retain only the most early record for each user_id from join.complete.df
join.complete.1st_time.df<-join.complete.df%>% group_by(user_id) %>%  filter(order_date == min(order_date)) # the filter function here make sure that each user_id only appeared once, and we select the row according to the earliesr order date of the uer_id

# create purchase activity data frame
activity_1storder_cohort_df = join.complete.1st_time.df %>% group_by(CohortGroup, weeks_after_signup) %>% summarise( FirstTimeUsers = n_distinct(user_id))
head(activity_1storder_cohort_df)

# join activity_1storder_cohort_df and base_cohort_df
FistTimeUsers_cohort_df = inner_join(activity_1storder_cohort_df, base_cohort_df, 
                  by = 'CohortGroup')

# Cohort table by the counts of users
FistTimeUsers_retention_df = FistTimeUsers_cohort_df %>% select(CohortGroup,  cohortname, TotalUsers, weeks_after_signup, FirstTimeUsers) %>%spread(weeks_after_signup, FirstTimeUsers)
head(FistTimeUsers_retention_df)
# change the column names 
for (colno in colnames(FistTimeUsers_retention_df)[-1:-3]){
  colnames(FistTimeUsers_retention_df)[as.numeric(colno)+3]<-paste0("Days",as.numeric(colno)*7-7,"to",as.numeric(colno)*7-1)
}

head(FistTimeUsers_retention_df)

# the percentage version of the 1st time user dataframe 
FistTimeUsers_retention.perc<-sapply(FistTimeUsers_retention_df[,-1:-3], function(x) paste0(round(x/FistTimeUsers_retention_df$TotalUsers*100,2),"% 1st time"))
FistTimeUsers_retention.perc<-gsub( "NA% 1st time", "",FistTimeUsers_retention.perc)

# bind the first time user count matrix and percentage  matrix together
FistTimeUsers_retention_df.perc<-paste.matrix(FistTimeUsers_retention.perc, "(", FistTimeUsers_retention_df[,-1:-3],")")
colnames(FistTimeUsers_retention_df.perc)<-colnames(FistTimeUsers_retention.perc)
FistTimeUsers_retention_df.perc<-cbind(FistTimeUsers_retention_df[,1:3],FistTimeUsers_retention_df.perc)
head(FistTimeUsers_retention_df.perc)
```

# Bind the total user cohort table and first time user cohort table
```{r}
# mask the first three columns of FistTimeUsers_retention_df.perc
FistTimeUsers_retention_df.perc.NA<-FistTimeUsers_retention_df.perc
FistTimeUsers_retention_df.perc.NA[,1:3]<-NA

# Make a for loop to alternatively input the rows from the total user cohort table (cohort_retention_df.perc) and the first time user cohort table (FistTimeUsers_retention_df.perc)
combined.matrix<-data.frame() # empty data frame for storing the data

for (i in 1:nrow(cohort_retention_df.perc)){
  rbined.matrix<-rbind(cohort_retention_df.perc[i,], FistTimeUsers_retention_df.perc.NA[i,])
  combined.matrix<-rbind(data.frame(combined.matrix),data.frame(rbined.matrix))
}

head(combined.matrix)

# save the data frame as csv
write.csv(combined.matrix, paste0(pathname, "cohort_analysis_output.csv"), row.names = FALSE)
```

# Data visualization
# Total user cohort heatmap
```{r}
# Below, I'm trying to make a heat map of the retention rate for total buyers 
# delete the characters and convert the data frame to numeric
cohort_retention.perc2<-sapply(data.frame(gsub("% orderers", "", cohort_retention.perc)),  function(x) as.numeric(x))
cohort_retention_df.perc2<-cbind(cohort_retention_df.perc[,1:3],cohort_retention.perc2)
head(cohort_retention_df.perc2)

# melt the dataframes for plotting
cohort_retention_df.perc.melt = gather(cohort_retention_df.perc2, "Order.date.since.signup", "BuyingUsersPerc", 4:ncol(cohort_retention_df.perc2))
cohort_retention_df.perc.melt<-cohort_retention_df.perc.melt[-which(is.na(cohort_retention_df.perc.melt)),]
cohort_retention_df.perc.melt[1:10,] # remove NAs
# reorder the Order.date.since.signup
cohort_retention_df.perc.melt$Order.date.since.signup<-factor(cohort_retention_df.perc.melt$Order.date.since.signup, levels = unique(cohort_retention_df.perc.melt$Order.date.since.signup))

# ggplot
cohort_retention.perc.ggplot<-ggplot(cohort_retention_df.perc.melt) +
  geom_raster(aes(x = Order.date.since.signup,
                  y = reorder(cohortname , desc(cohortname)),
                  fill = BuyingUsersPerc), show.legend=TRUE) +
    scale_fill_viridis_c(name = "Purchase rate (%)") + 
  geom_text(aes(x = Order.date.since.signup,
                y = reorder(cohortname, desc(cohortname)),
                label = BuyingUsersPerc), col = "white",size = 3) +
#  scale_y_discrete(labels = unique(cohort_retention_df.perc.melt$cohortname))+
  labs(
    title = "(A) Cohort analysis for weekly orders",
    x = "Days between purchase date and sign-up date",
    y = "Cohort Group"
  )+
  theme(axis.text.x=element_text(angle = -90, hjust = 0))

cohort_retention.perc.ggplot
```

## First Time user cohort heatmap
```{r}
# Below, I'm trying to make a heat map of the retention rate for first time buyers 
# delete the characters and convert the data frame to numeric
FistTimeUsers_retention.perc2<-sapply(data.frame(gsub("% 1st time", "", FistTimeUsers_retention.perc)),  function(x) as.numeric(x))
FistTimeUsers_retention_df.perc2<-cbind(FistTimeUsers_retention_df.perc[,1:3],FistTimeUsers_retention.perc2)
head(FistTimeUsers_retention_df.perc2)

# melt the dataframes for plotting
FistTimeUsers_retention_df.perc.melt = gather(FistTimeUsers_retention_df.perc2, "Order.date.since.signup", "BuyingUsersPerc", 4:ncol(FistTimeUsers_retention_df.perc2))
FistTimeUsers_retention_df.perc.melt<-FistTimeUsers_retention_df.perc.melt[-which(is.na(FistTimeUsers_retention_df.perc.melt)),]# remove NAs
FistTimeUsers_retention_df.perc.melt[1:10,]# Check out the first 10 rows
# reorder the Order.date.since.signup
FistTimeUsers_retention_df.perc.melt$Order.date.since.signup<-factor(FistTimeUsers_retention_df.perc.melt$Order.date.since.signup, levels = unique(FistTimeUsers_retention_df.perc.melt$Order.date.since.signup))

# ggplot
FistTimeUsers_retention.perc.ggplot<-ggplot(FistTimeUsers_retention_df.perc.melt) +
  geom_raster(aes(x = Order.date.since.signup,
                  y = reorder(cohortname , desc(cohortname)),
                  fill = BuyingUsersPerc)) +
    scale_fill_viridis_c(name = "Purchase rate (%)") + 
  geom_text(aes(x = Order.date.since.signup,
                y = reorder(cohortname, desc(cohortname)),
                label = BuyingUsersPerc), col = "white", size = 3) +
#  scale_y_discrete(labels = unique(FistTimeUsers_retention_df.perc.melt$cohortname))+
  labs(
    title = "(B) Cohort analysis for first-time orders",
    x = "Days between the first purchase and sign-up date",
    y = "Cohort Group"
  )+
  theme(axis.text.x=element_text(angle = -90, hjust = 0))

FistTimeUsers_retention.perc.ggplot
```

## combine the two heatmaps
```{r}
grid.arrange(cohort_retention.perc.ggplot, FistTimeUsers_retention.perc.ggplot)

# save heatmaps
tiff(paste0(pathname, "retention.heatmaps.tiff"), height = 15, width = 12, unit="in", res = 600)
grid.arrange(cohort_retention.perc.ggplot, FistTimeUsers_retention.perc.ggplot)
dev.off()
```

## Total user and first time user cohort line chart
We will use facet_wrap function to look at each cohort group
```{r}
# add additional column as week no.
weekno.df<-data.frame(Order.date.since.signup = unique(cohort_retention_df.perc.melt$Order.date.since.signup),
                      weekNo = 1:length(unique(cohort_retention_df.perc.melt$Order.date.since.signup)))

cohort_retention_df.perc.melt2<-left_join(cohort_retention_df.perc.melt, weekno.df)
head(cohort_retention_df.perc.melt2)

FistTimeUsers_retention_df.perc.melt2<-left_join(FistTimeUsers_retention_df.perc.melt, weekno.df)
head(FistTimeUsers_retention_df.perc.melt2)

# plotting line chart
cohort_retention.perc.linechart<-ggplot(cohort_retention_df.perc.melt2) +
  geom_line(aes(x = weekNo,y = BuyingUsersPerc, 
                col = cohortname)) +
  labs(
    title = "(A) Cohort analysis for weekly orders",
    x = "Number of weeks between purchase date and sign-up date",
    y = "Purchase rate (%)",
    col = "Cohort Group (user sign-up date)"
  )+
  theme_bw()

cohort_retention.perc.linechart # as you can see, it is hard to see each cohort one by one

# plotting line chart using facetwrap to show ,the cohort one by one
cohort_retention.perc.linechart.facet<-ggplot() +
  geom_line(aes(x = weekNo,y = BuyingUsersPerc), data = cohort_retention_df.perc.melt2)+ # use solid line for total users
  geom_line(aes(x = weekNo,y = BuyingUsersPerc), data = FistTimeUsers_retention_df.perc.melt2, linetype = "dotted") + # use dotted line for first time orders
  facet_wrap(~cohortname, ncol = 5)+
  labs(
    x = "Number of weeks between purchase date and sign-up date",
    y = "Purchase rate (%)",
    col = "Cohort Group (user sign-up date)"
  )+
  theme_bw()

cohort_retention.perc.linechart.facet
```